Static contract checking for Haskell

### 2011-12-19 paper/paper.tex
Paper: cleaned up the description of H' and the data-type translation.

The syntax of $\mathcal{H}'$ is defined in figure~\ref{hprime-stx}. A module
is a list of toplevel definitions, claims that functions statisfy
contracts and data definitions.

\begin{itemize}
\item There's no $\lambda$-abstraction, because we can always lift
  them to a toplevel declaration.
\item We do not allow nested case expressions, because once again, we
  can always lift them to the toplevel.
\item Until section~\ref{ho} we will only consider full application of
  functions ($\full{f}(x,y)$),\note[nc]{

Not true: the definition of satisfaction for arrow contracts
$x:c_1 \to c_2$ already includes partial applications.  We could add a
special cases for each function arity, e.g.
\begin{eqnarray*}
&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c} \\
&&      :=\forall x_1, \ldots, x_n. \min{\full{f}(x_1, \ldots, x_n)} \\
&&\quad   \to \left( \bigvee_{i=1}^n \ctrans{x_i \notin c_i} \right)
              \lor \ctrans{\full{f}(x_1, \ldots, x_n) \in c}
\end{eqnarray*}
for functions of arity $n$, to make the claim true, but that would add
more clutter.
}, in order to remove clutter. Dealing with
  partial application is not hard but a bit cumbersome.
\end{itemize}

@@

\begin{figure}
 \centering
  \[  \begin{array}{rcl}
    \phi &:=& \forall x.\phi \mid \text{\add[nc]{$\exists x.\phi \mid$}} \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t) \mid \min{t} \\
    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi
  \end{array} \]
  \caption{First-order logic syntax}
  \label{fol-stx}
\end{figure}

@@

\begin{itemize}
\item $(\phi_\text{injective})$ For each $K_i$ of arity $a_i$ we introduce selectors
  $\proj{k}{K_i}$, which are the projection of $K_i~x_1~\dots~x_{a_i}$
  on its $k$-th component, for $k = 1\text{ to }a_i$.
\item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free. \note[nc]{We use the contract
    translation for arrows, and this translation is principled.}
  \item $\to$: a value $t$ of type $T$\note[nc]{
    ``type'' is used loosely here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free. \note[nc]{XXX: This translation is ad-hoc.}

  Note that this is not true for functions: a function is not required
  to use all of its arguments. \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  being crash-free or not doesn't matter.

>>>>

The syntax of $\hprime$ is defined in figure~\ref{hprime-stx}. A
module is a list of toplevel definitions, claims that functions
statisfy contracts and data definitions.  The $\hprime$ language we
present is minimalistic, but the implementation supports a little
more.

\begin{itemize}
\item $\hprime$ doesn't include $\lambda$-abstractions, only top-level
  function definitions, because nested $\lambda$s can be lifted to the
  top-level.
\item $\hprime$ doesn't include nested case expressions, because
  nested case expressions can be lifted to the top-level.  The
  implementation actually supports nested case expressions.

\note[nc]{
Not true that we only consider full application of functions: the definition of satisfaction for arrow contracts
$x:c_1 \to c_2$ already includes partial applications.  We could add a
special cases for each function arity, e.g.
\begin{eqnarray*}
&&\ctrans{f \in x_1:c_1 \to \cdots \to x_n:c_n \to c}^v \\
&&      :=\forall x_1, \ldots, x_n. \min{\full{f}(x_1, \ldots, x_n)} \\
&&\quad   \to \left( \bigwedge_{i=1}^n \ctrans{x_i \in c_i}^{\bar{v}} \right)
              \to \ctrans{\full{f}(x_1, \ldots, x_n) \in c}^v
\end{eqnarray*}
for functions of arity $n$, to make the claim true, but that would add
more clutter.
}, in order to remove clutter. Dealing with
  partial application is not hard but a bit cumbersome.
\end{itemize}

@@

\begin{figure}
 \centering
  \[  \begin{array}{rcl}
    \phi &:=& \forall x.\phi \mid \exists x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t) \mid \min{t} \\
    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi
  \end{array} \]
  \caption{First-order logic syntax}
  \label{fol-stx}
\end{figure}

@@

\begin{itemize}
\item $(\phi_\text{injective})$ Constructors are injective.  That is, for
  each term constructor $K$,
  if $K e_1 \cdots e_n = K e'_1 \cdots e'_n$, then $e_i = e'_i$ for
  all $i$.  We axiomitize this via projections $\proj{i}{K}$: for each $i$,
  $\proj{i}{K}(K e_1 \cdots e_n) = e_i$.
\item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free.  For $K$ has arity $n$
    we assert $\ctrans{K \in \cfc^n \to \cfc}^-$.

  \item $\to$: a value $t$ of type $T$\note[nc]{
    ``type'' is used loosely here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free, since we can use
    a case expression to extract any packed value.
    \note[nc]{XXX: This translation is ad-hoc.}

  Note that this is not true for functions: a function is not required
  to use all of its arguments. \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  is irrelevant.
